---
output: html_document
---

```{r}
library(MASS)
```

2 type up 2 here


3 a) Normal equations: 
```{r}
data <- read.table("economic_data.txt", header=T)

M <- as.matrix(cbind(A0=rep(1,16),data[,2:7]))
Y <- data$B

#divide into smaller groups of alphas
alpha <- solve(t(M) %*% M, tol=10^-30) %*% (t(M) %*% Y)
```
The resulting alphas are `r alpha`.

b) Steepest Descent:
```{r}
#doing initial values for alpha1 - alpha6
initial <- rep(0, 7)

#the gradient is for the least squares error equation
grad.alpha <- function(alpha=rep(0,7)) {
  d.alphas <- 2*t(M)%*%M%*%alpha - 2*t(M)%*%Y
  return(as.vector(d.alphas))
}

norm.alpha <- function(x) {
  norm <- sqrt(sum(x^2))
  return(norm)
}

bisect.alpha <- function(alpha=rep(0,7), tol=.001) {
  s.a <- -10
  s.b <- 10
  s.c = (s.a + s.b) /2
  direction <- -grad.alpha(alpha)/norm.alpha(grad.alpha(alpha))
  
  
  h.a <- t(grad.alpha(alpha=alpha+s.a*direction))%*%direction
  h.b <- t(grad.alpha(alpha+s.b*direction))%*%direction
  h.c <- t(grad.alpha(alpha=alpha+s.c*direction))%*%direction
  
  while ((s.b-s.a)/2 > tol & abs(t(grad.alpha(alpha+s.c*direction))%*%direction) > tol) {
    grad.a <- t(grad.alpha(alpha=alpha+s.a*direction))%*%direction
    grad.c <- t(grad.alpha(alpha=alpha+s.c*direction))%*%direction
    
    ifelse(grad.a * grad.c < 0, s.b <- s.c, s.a <- s.c)
    s.c <- (s.a + s.b) /2  
  }
  return(s.c)
}

debug(bisect.alpha)
undebug(bisect.alpha)
bisect.alpha()

##graph to show we're finding the minimum
s.a <- seq(-10,10,.01)
alpha <- rep(5,7)
direction <- -grad.alpha(alpha)/norm(grad.alpha(alpha))
y <- sapply(1:2001, function(x) t(Y-M%*%(alpha+s.a[x]*direction))%*%(Y-M%*%(alpha+s.a[x]*direction)))
y.g <- sapply(1:2001, function(x) t(grad.alpha(alpha=alpha+s.a[x]*direction))%*%direction)

#h(s)
qplot(x=s.a, y=y)

#g(s)
qplot(x=s.a, y=y.g)

#runs, but not very accurate
#takes a long time, ~hour to get to final result
#run once would all same alphas (current version)
#run second time with large alphas
#run third with varying size alphas
descent.alpha <- function(initial = rep(1,7), target = 1) {
  alpha.new <- initial
  path <- rbind(rep(0,7),alpha.new)
  gradients <- NULL
  i <- 2
  
  while(abs(norm.alpha(grad.alpha(path[i,]))-norm.alpha(grad.alpha(path[i-1,]))) > 0 && norm.alpha(grad.alpha(alpha.new)) > target) {
    gradients <- rbind(gradients,norm.alpha(grad.alpha(alpha.new)))
    alpha.old <- alpha.new
    
    dir <- -grad.alpha(alpha.old) / norm.alpha(grad.alpha(alpha.old))
    step <- bisect.alpha(alpha=alpha.old, tol=.001)
    alpha.new <- alpha.old + dir * step
    path <- rbind(path,alpha.new)
    print(norm.alpha(grad.alpha(alpha.new)))
    write.table(as.data.frame(alpha.new), file="alphas.txt", append=TRUE)
    i <- i + 1
  }
  return(alpha.new)
}

descent.alpha()

```

c)
```{r}
attach(data)
lm(B ~ A1 + A2 + A3 + A4 + A5 + A6)
```

4. Banana function
The minimum of f is (1, 1) (why?). Starting at the point (4, 4) use Steepest descent with a
bisection root finding method to try and ”find” the minimum. Produce a plot of the path
your iteration takes, i.e. plot x
(0), x(1), x(2), . . . . What does the path look like? You may just
want to plot a few consecutive x(i) values if the points are too dense to tell apart.

```{r}
#doing initial values for x,y
initial <- c(4,4)

b.fxn <- function(params = initial) {
  f.xy <- 100*(params[2]-params[1]^2)^2 + (1-params[1])^2
  return(f.xy)
}

grad.bf <- function(params = c(2,2)) {
  d.x <- -400*(params[2]-params[1]^2)*params[1]-2*(1-params[1])
  d.y <- 200*(params[2]-params[1]^2)
  return(c(d.x,d.y))
}

norm.bf <- function(x) {
  norm <- sqrt(sum(x)^2)
  return(norm)
}

#need to resolve: choosing initial s.a and s.b, what to return
bisect.bf <- function(xy=c(2,2), tol=.001, s.a=-100, s.b=100) {
  s.c = (s.a + s.b) /2
  
  direction <- grad.bf(params=xy)/norm.bf(grad.bf(params=xy))
  
  while ((s.b-s.a)/2 > tol && abs(t(grad.bf(params=xy + direction*s.c))%*%direction) > tol) {
    grad.a <- t(grad.bf(params=(xy+s.a*direction)))%*%direction
    grad.c <- t(grad.bf(params=(xy+s.c*direction)))%*%direction
    
    if(grad.a * grad.c < 0) {
      s.b <- s.c
    } else {
      s.a <- s.c
    }
    s.c <- (s.a + s.b) /2  
  }
  return(s.c)
}

debug(bisect.bf)
undebug(bisect.bf)
bisect.bf()


#figure out what it is going way off course
#improve print out to file
descent.bf <- function(initial = c(2,2), target = .1) {
  xy.new <- initial
  path <- NULL
  gradients <- NULL
  
  while(norm.bf(grad.bf(params=xy.new)) > target ) {
    path <- rbind(path,xy.new)
    gradients <- rbind(gradients,norm.bf(grad.bf(xy.new)))
    xy.old <- xy.new
    
    dir <- -grad.bf(xy.old) / norm.bf(grad.bf(xy.old))
    step <- bisect.bf(xy=xy.old)
    xy.new <- xy.old + dir * step
    write.table(xy.new, "banana.txt", append=TRUE)
    print(xy.new)
    
  }
  return(rbind(path,xy.new))
}

descent.bf()

```




