---
output: html_document
---

```{r}
library(MASS)
```


3 a) Normal equations: 
```{r}
data <- read.table("economic_data.txt", header=T)

M <- as.matrix(cbind(A0=rep(1,16),data[,2:7]))
Y <- data$B

alpha <- solve(t(M) %*% M, tol=10^-30) %*% (t(M) %*% Y)
```
The resulting alphas are `r alpha`.

b) Steepest Descent:
```{r}
#doing initial values for alpha1 - alpha6
initial <- rep(0, 7)

#the gradient is for the least squares error equation
#so then L(alpha) = (y-M*alpha)(y-M*alpha)
#L'(alpha)= -1*(y-M*alpha) * (-1)*(y-M*alpha) = (y-M*alpha)(y-M*alpha)
grad.alpha <- function(alpha=rep(0,7)) {
  d.alphas <- 2*t(M)%*%M%*%alpha - 2*t(M)%*%Y
  return(as.vector(d.alphas) )
}

norm.alpha <- function(x) {
  norm <- sqrt(sum(x^2))
  return(norm)
}

bisect.alpha <- function(alpha=rep(0,7), tol=.001) {
  s.a <- 0
  s.b <- 100
  s.c = (s.a + s.b) /2
  direction <- -grad.alpha(alpha)/norm(grad.alpha(alpha))
  
  
  h.a <- t(grad.alpha(alpha=alpha+s.a*direction))%*%direction
  h.b <- t(grad.alpha(alpha+s.b*direction))%*%direction
  h.c <- t(grad.alpha(alpha=alpha+s.c*direction))%*%direction
  
  while ((s.b-s.a)/2 > tol & abs(t(grad.alpha(alpha+s.c*direction))%*%direction) > tol) {
    grad.a <- t(grad.alpha(alpha=alpha+s.a*direction))%*%direction
    grad.c <- t(grad.alpha(alpha=alpha+s.c*direction))%*%direction
    
    ifelse(grad.a * grad.c < 0, s.b <- s.c, s.a <- s.c)
    s.c <- (s.a + s.b) /2  
  }
  return(s.c)
}

debug(bisect.alpha)
undebug(bisect.alpha)
bisect.alpha()

##graph to show we're finding the minimum
s.a <- seq(-10,10,.01)
alpha <- rep(5,7)
direction <- -grad.alpha(alpha)/norm(grad.alpha(alpha))
y <- sapply(1:2001, function(x) t(Y-M%*%(alpha+s.a[x]*direction))%*%(Y-M%*%(alpha+s.a[x]*direction)))
y.g <- sapply(1:2001, function(x) t(grad.alpha(alpha=alpha+s.a[x]*direction))%*%direction)

#h(s)
qplot(x=s.a, y=y)

#g(s)
qplot(x=s.a, y=y.g)

#edit this to fit once bisect function works
descent.alpha <- function(initial = rep(1,7), target = .1) {
  alpha.new <- initial
  path <- rbind(rep(0,7),alpha.new)
  gradients <- NULL
  i <- 2
  
  while(norm(grad.alpha(path[i,]))-norm(grad.alpha(path[i-1,])) > target ) {
    path <- rbind(path,alpha.new)
    gradients <- rbind(gradients,norm(grad.alpha(alpha.new)))
    alpha.old <- alpha.new
    
    dir <- -grad.alpha(alpha.old) / norm(grad.alpha(alpha.old))
    step <- bisect.alpha(alpha=alpha.old, tol=.001)
    alpha.new <- alpha.old + dir * step
    print(norm(grad.alpha(alpha.new)))
    write.table(as.data.frame(alpha.new), file="alphas.txt", append=TRUE)
  }
  return(rbind(path,alpha.new))
}

descent.alpha()

```

c)
```{r}
attach(data)
lm(B ~ A1 + A2 + A3 + A4 + A5 + A6)
```

4. Banana function
The minimum of f is (1, 1) (why?). Starting at the point (4, 4) use Steepest descent with a
bisection root finding method to try and ”find” the minimum. Produce a plot of the path
your iteration takes, i.e. plot x
(0), x(1), x(2), . . . . What does the path look like? You may just
want to plot a few consecutive x(i) values if the points are too dense to tell apart.

```{r}
#doing initial values for x,y
initial <- c(4,4)

b.fxn <- function(params = initial) {
  f.xy <- 100*(params[2]-params[1]^2)^2 + (1-params[1])^2
  return(f.xy)
}


##f(a)*f(b) need to be less than 0, this is happening yet
bisect.bf <- function(alphas=c(0,0), tol=.001, direction = .01, s.a=-1, s.b=1) {
  s.c = (s.a + s.b) /2
  
  while (s.c > tol & b.fxn(params=alphas) != 0) {
    grad.a <- grad.bf(params=(alphas+s.a*direction))*direction
    grad.c <- grad.bf(params=(alphas+s.c*direction))*direction
    
    if(grad.a[1] * grad.c[1] < 0 | grad.a[2] * grad.c[2] < 0) {
      s.b <- s.c
    } else {
      s.a <- s.c
    }
    s.c <- (s.a + s.b) /2  
  }
  return(c(s.a,s.b,s.c))
}

debug(bisect.bf)
undebug(bisect.bf)
bisect.bf()

grad.bf <- function(params = initial) {
  d.x <- -400*params[1]*params[2] + 400*params[1]^3 - 2 + 2*params[1]
  d.y <- 200*(params[2]-params[1]^2)
  return(c(d.x,d.y))
}

norm.bf <- function(x) {
  norm <- sqrt(sum(x^2))
  return(norm)
}

descent <- function(initial = c(2,2), target = .1) {
  xy.new <- initial
  path <- NULL
  gradients <- NULL
  
  while(norm(grad(params=xy.new)) > target ) {
    path <- rbind(path,xy.new)
    gradients <- rbind(gradients,norm(grad(xy.new)))
    xy.old <- xy.new
    
    dir <- -grad(xy.old) / norm(grad(xy.old))
    step <- 
    xy.new <- xy.old + dir * step
    print(norm(grad(xy.new)))
  }
  return(rbind(path,xy.new))
}

```




