---
output:
  html_document:
    fig_width: 5
---
<p>Elaine Ayo</p>
<p>Math 504</p>
<p>Jan. 25, 2015</p>

```{r echo=FALSE}
library(MASS)
library(ggplot2)
library(scatterplot3d)
library(reshape2) 
```

<p>2 a)</p>
<p>Graph of function:</p>
```{r, echo=FALSE, error=FALSE, warning=FALSE}
f.xy <- function(params=data.frame(x=0, y=0)) {
  params <- params
  xy <- 2000*(params$x)^2 + (params$y)^2
  return(xy)
}

x.s <- seq(-10,10,1)
y.s <- x.s

z <- sapply(1:21, function(i) sapply(1:21, function(j) f.xy(params=data.frame(x=x.s[i],y=y.s[j]))))

base <- persp(x.s, y.s, z, theta = 60, phi = 20, expand = 0.5, col = "lightblue", ltheta = 120, shade = 0.75, border=NA, xlab = "X", ylab = "Y", zlab = "Z") 
base + title("f(x,y) with x,y = [-10,10]")
```

<p>Steepest descent function: </p>
```{r, eval=FALSE}
grad <- function(params=data.frame(x=10, y=10)) {
  d.x <- 4000*params$x
  d.y <- 2*params$y
  return(c(d.x,d.y))
}

norm <- function(x) {
  norm <- sqrt(sum(x)^2)
  return(norm)
}

descent <- function(initial = data.frame(x=10, y=10), target = 1) {
  #initialize all the things
  xy.new <- data.frame(x=10, y=10)
  grad.new <- grad(params=xy.new)
  
  while(norm(grad.new) > target ) {
    #making room for new values
    xy.old <- xy.new
    grad.old <- grad.new
    
    dir <- -grad.old / norm(grad.old)
    step <- 1
    xy.new <- xy.old + dir * step
    grad.new <- grad(params=xy.new)
    
    results <- data.frame(x=xy.new$x, y=xy.new$y, norm.grad=norm(grad.new))
    write.table(results, "2a4.csv", row.names = FALSE, col.names = FALSE, append=TRUE)
  }
  return(xy.new)
}

xy <- descent()
```

<p>Trial 1 (cyan): Step size: .1; target: .1; No. of steps: 582; Roots: x=0.0003217926 / y=-0.6003218 </p>
<p>Trial 2 (pink): Step size: .01; target: .01; No. of steps: 5,674; Roots: x=-1.003020e-06 / y=1.003020e-06</p>
<p>Trial 3 (green): Step size: .001; target: .001; No. of steps: 20,000; Roots: x=0 / y=3.331363e-14</p>
<p>Trial 4 (Gray): Step size: 1; target: 1; No. of steps: 3,296; Roots: x=-5.809212e-05 / y=5.809212e-05</p>

```{r}

data1 <- read.table("2a.csv", header=FALSE)
x.path1 <- data1$V1
y.path1 <- data1$V2

data2 <- read.table("2a2.csv", header=FALSE)
x.path2 <- data2$V1
y.path2 <- data2$V2

data3 <- read.table("2a3.csv", header=FALSE)
x.path3 <- data3$V1
y.path3 <- data3$V2

data4 <- read.table("2a4.csv", header=FALSE)
x.path4 <- data4$V1
y.path4 <- data4$V2

base <- persp(x.s, y.s, z, theta = 60, phi = 20, expand = 0.5, col = "lightblue", ltheta = 120, shade = 0.75, border=NA, xlab = "X", ylab = "Y", zlab = "Z") + title("f(x,y) with various paths to minimum")

lines(trans3d(x=x.path4, y = y.path4, z = f.xy(params=data.frame(x.path4,y.path4)), pmat = base), col = 8, lwd = 4)

lines(trans3d(x=x.path1, y = y.path1, z = f.xy(params=data.frame(x.path1,y.path1)), pmat = base), col = 5, lwd = 2)

lines(trans3d(x=x.path2, y = y.path2, z = f.xy(params=data.frame(x.path2,y.path2)), pmat = base), col = 6, lwd = 1)

lines(trans3d(x=x.path3+.1, y = y.path3+.1, z = f.xy(params=data.frame(x.path3,y.path3))+.1, pmat = base), col = 3, lwd = .2)
```
<p>b)</p>
<p>Bisection function:</p>
```{r}
bisect <- function(xy=data.frame(x=10, y=10), tol=.001, s.a=0, s.b=100) {
  xy=data.frame(x=10, y=10)
  s.a <- 0
  s.b <- 100
  s.c <- (s.a + s.b) /2
  direction <- -grad(params=xy)/norm(grad(params=xy))
  g.c <- t(grad(params=(xy+s.c*direction)))%*%direction
  
  while ((s.b-s.a)/2 > tol && abs(g.c) > tol) {
    g.a <- t(grad(params=(xy+s.a*direction)))%*%direction
    g.c <- t(grad(params=(xy+s.c*direction)))%*%direction
    
    ifelse(g.a * g.c < 0, s.b <- s.c, s.a <- s.c)
    s.c <- (s.a + s.b) /2  
    g.c <- t(grad(params=(xy+s.c*direction)))%*%direction
  }
  return(s.c)
}

bisect1 <- bisect()
```

<p>The function finds a minimum at `r bisect1`, which is confirmed as the root by the graph below.</p>

```{r, echo=FALSE, fig.align='center'}
s.a <- seq(-30,30,1)
xy <- data.frame(x=10,y=10)
direction <- -grad(xy)/norm(grad(xy))
g.s <- sapply(1:61, function(x) t(grad(params=(xy+s.a[x]*direction)))%*%direction)

#h(s)
qplot(x=s.a, y=g.s, geom="line") + ggtitle("g(s) for f(x,y) when x=10 and y=10")
```

<p>Steepest descent function: </p>
```{r, eval=FALSE}

##something weird is going on, graph 2a5 and see what's up
#compare to how gradients graphed for consistent step size runs

descent.b <- function(initial = data.frame(x=10, y=10), target = 1) {
  #initialize all the things
  xy.new <- data.frame(x=10, y=10)
  grad.new <- grad(params=xy.new)
  
  while(norm(grad.new) > target ) {
    #making room for new values
    xy.old <- xy.new
    grad.old <- grad.new
    
    dir <- -grad.old / norm(grad.old)
    step <- bisect(xy=xy.old)
    xy.new <- xy.old + dir * step
    grad.new <- grad(params=xy.new)
    
    results <- data.frame(x=xy.new$x, y=xy.new$y, norm.grad=norm(grad.new))
    write.table(results, "2a5.csv", row.names = FALSE, col.names = FALSE, append=TRUE)
  }
  return(xy.new)
}

xy <- descent.b()
```

<p>c)</p>

<p>3. a)</p>
```{r}
heights <- read.table("children_heights", header=TRUE)

#group1: boys_2  
#group2: boys_9
#group3: boys_18

#put this in a list and write function to dynamically create list of means for given N, same logic for making multiple plots
#mean1 <- mean(heights$Boys_2, na.rm=TRUE)
#mean2 <- mean(heights$Boys_9, na.rm=TRUE)
#mean3 <- mean(heights$Boys_18, na.rm=TRUE)

means <- sapply(1:3, function(x) mean(heights[,x], na.rm=TRUE))

#N=78 (all boys in data, minus 3 is 75)
#var.boys <- (1/75) * (sum((heights$Boys_2 - mean1)^2, na.rm=TRUE) + sum((heights$Boys_9 - mean2)^2, na.rm=TRUE) + sum((heights$Boys_18 - mean3)^2, na.rm=TRUE))

vars <- (1/75) * sum(sapply(1:3, function(x) sum((heights[,x]-means[x])^2, na.rm=TRUE)))

#boys <- c(heights[,1], heights[,2], heights[,3])
#boys <- boys[!is.na(boys)]

boys.melt <- melt(heights[,1:3], na.rm= TRUE)
colnames(boys.melt) <- c("Class","Height")

class.boys <- function(boy=90) {
  means <- sapply(1:3, function(x) mean(heights[,x], na.rm=TRUE))

  var.boys <- (1/75) * sum(sapply(1:3, function(x) sum((heights[,x]-means[x])^2, na.rm=TRUE)))
  
  #class <- data.frame(g1 = dnorm(boy,mean=mean[1],sd=sqrt(var.boys)), g2 = dnorm(boy,mean=mean[2],sd=sqrt(var.boys)), g3 = dnorm(boy,mean=mean[3],sd=sqrt(var.boys)))
  
  class <- sapply(1:3, function(x) dnorm(boy, mean=means[x], sd=sqrt(var.boys)))

  return(which.max(class))
}

classes.boy <- factor(sapply(1:78, function(x) class.boys(boy=boys.melt$Height[x])), labels=c("Boys_2", "Boys_9", "Boys_18"))


final.boys <- data.frame(Height=boys, Original=factor(boys.melt$Class), New=classes.boy)

fails <- sum(ifelse(final.boys$Original == final.boys$New, 0, 1))

#All boys classified correctly
```
<p>All boys are classified correctly.</p>

<p>b)</p>
```{r}
heights <- read.table("children_heights", header=TRUE)

#group1: boys_2  
#group2: boys_9
#group3: boys_18
#group4: Girls_2  
#group5: Girls_9
#group6: Girls_18

#put this in a list and write function to dynamically create list of means for given N, same logic for making multiple plots
mean1 <- mean(heights$Boys_2, na.rm=TRUE)
mean2 <- mean(heights$Boys_9, na.rm=TRUE)
mean3 <- mean(heights$Boys_18, na.rm=TRUE)
mean4 <- mean(heights$Girls_2, na.rm=TRUE)
mean5 <- mean(heights$Girls_9, na.rm=TRUE)
mean6 <- mean(heights$Girls_18, na.rm=TRUE)

#N=78 boys + 96 girls - 6 classes = 168
var.kids <- (1/168) * (sum((heights$Boys_2 - mean1)^2, na.rm=TRUE) + sum((heights$Boys_9 - mean2)^2, na.rm=TRUE) + sum((heights$Boys_18 - mean3)^2, na.rm=TRUE) + sum((heights$Girls_2 - mean1)^2, na.rm=TRUE) + sum((heights$Girls_9 - mean2)^2, na.rm=TRUE) + sum((heights$Girls_18 - mean3)^2, na.rm=TRUE))


kids.melt <- melt(heights, na.rm= TRUE)
colnames(kids.melt) <- c("Class","Height")

class.kids <- function(kid=90) {
  mean1 <- mean(heights$Boys_2, na.rm=TRUE)
  mean2 <- mean(heights$Boys_9, na.rm=TRUE)
  mean3 <- mean(heights$Boys_18, na.rm=TRUE)
  mean4 <- mean(heights$Girls_2, na.rm=TRUE)
  mean5 <- mean(heights$Girls_9, na.rm=TRUE)
  mean6 <- mean(heights$Girls_18, na.rm=TRUE)

  #N=78 boys + 96 girls - 6 classes = 168
  var.kids <- (1/168) * (sum((heights$Boys_2 - mean1)^2, na.rm=TRUE) + sum((heights$Boys_9 - mean2)^2, na.rm=TRUE) + sum((heights$Boys_18 - mean3)^2, na.rm=TRUE) + sum((heights$Girls_2 - mean1)^2, na.rm=TRUE) + sum((heights$Girls_9 - mean2)^2, na.rm=TRUE) + sum((heights$Girls_18 - mean3)^2, na.rm=TRUE))
  
  class <- data.frame(g1 = dnorm(kid,mean=mean1,sd=sqrt(var.kids)), g2 = dnorm(kid,mean=mean2,sd=sqrt(var.kids)), g3 = dnorm(kid,mean=mean3,sd=sqrt(var.kids)), g4 = dnorm(kid,mean=mean4,sd=sqrt(var.kids)), g5 = dnorm(kid,mean=mean5,sd=sqrt(var.kids)), g6 = dnorm(kid,mean=mean6,sd=sqrt(var.kids)))

  return(as.numeric(which.max(class)))
}


classes.kids <- sapply(1:174, function(x) class.kids(kid=kids.melt$Height[x]))

final.kids <- data.frame(Height=kids.melt$Height, Original=c(rep(1:3,each=26),rep(4:6,each=32)), New=classes.kids)

fails <- sum(ifelse(final.kids$Original == final.kids$New, 0, 1))

#calculate if classified incorrectly, which two were messed up
```

<p>66/174, or about .3793 children were classified incorrectly, which was to be expected due to the overlapping age groups. In part a, there were very clean divisions between the height of any child at 2, 9, and 18 years of age. However, with the introduction of a second gender, it became trickier to cleanly classify the data on the basis of both gender and age.</p>

<p>c)</p>
```{r}
heights <- read.table("children_heights", header=TRUE)

#group1: boys_2  
#group2: boys_9
#group3: boys_18
#group4: Girls_2  
#group5: Girls_9
#group6: Girls_18

#put this in a list and write function to dynamically create list of means for given N, same logic for making multiple plots
mean1 <- mean(heights$Boys_2, na.rm=TRUE)
mean2 <- mean(heights$Boys_9, na.rm=TRUE)
mean3 <- mean(heights$Boys_18, na.rm=TRUE)
mean4 <- mean(heights$Girls_2, na.rm=TRUE)
mean5 <- mean(heights$Girls_9, na.rm=TRUE)
mean6 <- mean(heights$Girls_18, na.rm=TRUE)

#N=78 boys + 96 girls - 6 classes = 168
var.kids <- (1/168) * (sum((heights$Boys_2 - mean1)^2, na.rm=TRUE) + sum((heights$Boys_9 - mean2)^2, na.rm=TRUE) + sum((heights$Boys_18 - mean3)^2, na.rm=TRUE) + sum((heights$Girls_2 - mean1)^2, na.rm=TRUE) + sum((heights$Girls_9 - mean2)^2, na.rm=TRUE) + sum((heights$Girls_18 - mean3)^2, na.rm=TRUE))


kids.melt <- melt(heights, na.rm= TRUE)
colnames(kids.melt) <- c("Class","Height")

class.kids.sd <- function(kid=90) {
  mean1 <- mean(heights$Boys_2, na.rm=TRUE)
  mean2 <- mean(heights$Boys_9, na.rm=TRUE)
  mean3 <- mean(heights$Boys_18, na.rm=TRUE)
  mean4 <- mean(heights$Girls_2, na.rm=TRUE)
  mean5 <- mean(heights$Girls_9, na.rm=TRUE)
  mean6 <- mean(heights$Girls_18, na.rm=TRUE)

  #N=78 boys + 96 girls - 6 classes = 168
  sd1 <- sd(heights$Boys_2, na.rm=TRUE)
  sd2 <- sd(heights$Boys_9, na.rm=TRUE)
  sd3 <- sd(heights$Boys_18, na.rm=TRUE)
  sd4 <- sd(heights$Girls_2, na.rm=TRUE) 
  sd5 <- sd(heights$Girls_9, na.rm=TRUE)
  sd6 <- sd(heights$Girls_18, na.rm=TRUE)
  
  class <- data.frame(g1 = dnorm(kid,mean=mean1,sd=sd1), g2 = dnorm(kid,mean=mean2,sd=sd2), g3 = dnorm(kid,mean=mean3,sd=sd3), g4 = dnorm(kid,mean=mean4,sd=sd4), g5 = dnorm(kid,mean=mean5,sd=sd5), g6 = dnorm(kid,mean=mean6,sd=sd6))

  return(as.numeric(which.max(class)))
}


classes.kids.sd <- sapply(1:174, function(x) class.kids.sd(kid=kids.melt$Height[x]))

final.kids.sd <- data.frame(Height=kids.melt$Height, Original=c(rep(1:3,each=26),rep(4:6,each=32)), New=classes.kids.sd)

fails.sd <- sum(ifelse(final.kids.sd$Original == final.kids.sd$New, 0, 1))
```

<p>Adding the individual sigmas for the groups slightly improved the classifications, with 62/178 (0.3483) being classified correctly.</p>

<p>4 a)</p>

```{r}
f.x <- function(x,a) {
  f.x <- x^2 - a
  return(f.x)
}

df.x <- function(x) {
  df.x <- 2*x
  return(df.x)
}

mySqrt.a <- function(a=4, tol=.01) {
  #s.old needs to be larger than the root of a to work
  #since we don't know a, we can't set a constant or an equation, because a can be anywhere
  #from near zero to near infinity, the next best thing is s.old = a if a > 1 and
  #s.old = 1 if a < 1
  #because we know for sure a > sqrt(a) if a > 1, and if a < 1, a = 1 should be larger than the root
  #and a must be positive, otherwise the function goes marching off toward negative infinity because the original f.x doesn't get smaller
  if (a < 0) {
    return("a must be >= 0")
  } else {
    
  s.old <- ifelse(a > 1, a, 1)
  while(f.x(s.old, a) > tol) {
    s.new <- s.old - f.x(s.old,a)/df.x(s.old)
    s.old <- s.new
  }
  return(s.old)
  }
}

y <- -.2
mySqrt.a(a=y, tol=.0001)
sqrt(x)
```

<p>b) </p>
```{r}
f.x <- function(x,a) {
  f.x <- x^2 - a
  return(f.x)
}

df.x <- function(x) {
  df.x <- 2*x
  return(df.x)
}

mySqrt.b <- function(a=4, tol=.01) {
  if (a < 0) {
    return("a must be >= 0")
  } else {
    
  s.old <- ifelse(a > 1, a, 1)
  while(f.x(s.old, a) > tol) {
    #can we use square root function in taylor expansion?
    s.new1 <- s.old - df.x(s.old)/2 + sqrt(df.x(s.old)^2 - 2*2*f.x(s.old,a))/2
    s.new2 <- s.old - df.x(s.old)/2 - sqrt(df.x(s.old)^2 - 2*2*f.x(s.old,a))/2
    s.old <- max(s.new1, s.new2, na.rm=TRUE)
  }
  return(s.old)
  }
}

y <- .04
debug(mySqrt.b)
mySqrt.b(a=y, tol=.1)

sqrt(x)
```

<p>c) </p>

```{r}
f.x <- function(x,a) {
  f.x <- x^2 - a
  return(f.x)
}

df.x <- function(x) {
  df.x <- 2*x
  return(df.x)
}

mySqrt.c <- function(a=4) {
  if (a < 0) {
    return("a must be >= 0")
  } else {
  s.old <- ifelse(a > 1, a, 1)
  root <- uniroot(function(x) x^2 - a, c(0, s.old), tol = 0.0000001)$root
  return(root)
  }
}

y <- .05
mySqrt.c(a=y)
sqrt(y)
```

<p>d) Compare speeds:</p>
```{r}


```