---
output:
  html_document:
    fig_width: 5
---
Elaine Ayo

Math 504 | Homework 7

March 1, 2015

```{r echo=FALSE, message=FALSE}
library(MASS)
library(ggplot2)
library(scatterplot3d)
library(reshape2) 
library(pryr)
library(microbenchmark)
library(ElemStatLearn)
library(glmnet)

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

2. 
```{r}
#cleaning data
LifeExpectancy2003 <- read.csv("~/Dropbox/numericalmethods/LifeExpectancy2003.csv")
life <- LifeExpectancy2003[1:101,1:2]
```

a. Calculating $L(t)$.

So in the example given in the problem, $L(2) = P(lives to be 42) = P(not die between 40-41) \times P(not die between 41-42) = (1 - P(die between 40-41)) \times (1 - P(die between 41-42))$. This is because we know the person has lived to be 40, when they bought the policy, unless this is some sort of elaborate fraud. So to calculate the probability of them not dying, we take the complement of the probability of them dying during each of the given time periods.

Function to computer $L(t) \forall t=0,1, 2 ...$. using spreadsheet data. We can start at the 41st row, or the probability of dying between 40-41, because we know the person is already 40 years old at the start of the problem. I cut it off at t=85, because the oldest person to have ever lived was 122, and so even taking into account advances in medical science, anything much more than that seems to be getting a bit silly, especially since the data just assigns a blanket "1" to all ages over 100.

```{r}
#max age = 125
t <- c(1:85)
data <- life
l.t <- ifelse(t <= 61, cumprod(1-data$q.x[40+t]), 0)

#now calculating the spline function and graphing the probability of not dying each day between 40 and 125
life.fxn <- splinefun(t, l.t, method="natural")
x <- seq(1,85,1/12)
y <- life.fxn(x=x)
x <- x + 40

qplot(c(40,x),c(1,y), geom="line") + xlab("Age") + ylab("Probability of Not Dying") + ggtitle("Max Age of 125")
```

b. $L(i/12)$ is the likelihood the person is alive each month, m, after age 40. So it makes sense when calculating the expected value of the policy, we incorporate the probability the person is still alive at a given age in months over the potential span of time they would be making payments for the policy.

The average present value of the payments is $39,372.98, assuming N = 125. Interestingly, the difference between fitting with a max age of 125 or 140 was less than a dollar off from the value of fitting with N = 101, when the data shifts to probability of dying of 1 for all older ages.

```{r}
i <- seq(1,1020,1)
l.i <- life.fxn(x=i/12)
e.pv <- 200 * l.i * exp(-.05*(i/12) )
sum(e.pv)
```

3. a.
$$M = \begin{bmatrix} 
& 1 & x & x^2 & x^3 & -1 & -x & -x^2 & -x^3 & 0 & 0 & 0 & 0 & \\
& 0 & 0 & 0 & 0 & 1 & x & x^2 & x^3 & -1 & -x & -x^2 & -x^3 & \\
& 0 & 1 & 2x & 3x^2 & 0 & -1 & -2x & -3x^2 & 0 & 0 & 0 & 0 & \\
& 0 & 0 & 0 & 0 & 0 & 1 & 2x & 3x^2 & 0 & -1 & -2x & -3x^2 & \\
& 0 & 0 & 2 & 6x & 0 & 0 & -2 & -6x & 0 & 0 & 0 & 0 & \\
& 0 & 0 & 0 & 0 & 0 & 0 & 2 & 6x & 0 & 0 & -2x & -6x & \\
\end{bmatrix}$$

Since the entries of M are linearly independent, $dim(null(M)) = col(M) - row(M) = 12 - 6 = 6$.

```{r}
bones <- read.table("~/Dropbox/numericalmethods/BoneMassData.txt", header=TRUE, quote="\"")
male <- bones[which(bones$gender == 'male'), ]
female <- bones[which(bones$gender == 'female'), ]

M <- matrix(0, nrow = 6, ncol = 12)

f.x <- function(x) return(c(1, x, x^2, x^3))
df.x <- function(x) return(c(0, 1, 2*x, 3*x^2))   
d2f.x <- function(x) return(c(0, 0, 2, 6*x))

M[1, 1:4] <- f.x(x=15)
M[1, 5:8] <- -f.x(x=15)
M[2, 1:4] <- df.x(x=15)
M[2, 5:8] <- -df.x(x=15)
M[3, 1:4] <- d2f.x(x=15)
M[3, 5:8] <- -d2f.x(x=15)

M[4, 5:8] <- f.x(x=20)
M[4, 9:12] <- -f.x(x=20)
M[5, 5:8] <- df.x(x=20)
M[5, 9:12] <- -df.x(x=20)
M[6, 5:8] <- d2f.x(x=20)
M[6, 9:12] <- -d2f.x(x=20)

M
```

b. i.So we've created $\mathcal{F} where \mathcal{F} : { z \in \mathbb{R}^12 | Mz = 0}$ and we know $dim(\mathcal{F}) = dim(z \in \mathbb{R}^12 | Mz = 0) = dim(null(M)) = 6$. Given this, any function $f \in \mathcal{F}$ can be expressed as $f(x) = \Sigma_{i=1}^6 \alpha_i h_i(x)$ where $h_i(x)$ is one of the six basis functions $h_1(x) - h_6(x)$. What this essentially does is instead of the familar $\Sigma_{i=1}^M\Sigma_{j=1}^N \alpha_i x_j$, we have a set of functions that change which $alpha$ are used depending on which part of the data we are in.

ii. A. Given: $z = \begin{bmatrix} & a_0 & b_0 & c_0 & d_0 & a_1 & b_1 & c_1 & d_1 & a_2 & b_2 & c_2 & d_2 & \end{bmatrix}$
 and since:
$$M = \begin{bmatrix} 
& 1 & x & x^2 & x^3 & -1 & -x & -x^2 & -x^3 & 0 & 0 & 0 & 0 & \\
& 0 & 0 & 0 & 0 & 1 & x & x^2 & x^3 & -1 & -x & -x^2 & -x^3 & \\
& 0 & 1 & 2x & 3x^2 & 0 & -1 & -2x & -3x^2 & 0 & 0 & 0 & 0 & \\
& 0 & 0 & 0 & 0 & 0 & 1 & 2x & 3x^2 & 0 & -1 & -2x & -3x^2 & \\
& 0 & 0 & 2 & 6x & 0 & 0 & -2 & -6x & 0 & 0 & 0 & 0 & \\
& 0 & 0 & 0 & 0 & 0 & 0 & 2 & 6x & 0 & 0 & -2x & -6x & \\
\end{bmatrix}$$

$$M^T = \begin{bmatrix} 
& 1   & 0   & 0    & 0    & 0  & 0 & \\
& x   & 0   & 1    & 0    & 0  & 0 & \\
& x^2 & 0   & 2x   & 0    & 2  & 0 & \\
& x^3 & 0   & 3x^2 & 0    & 6x & 0 & \\
& -1   & 1   & 0    & 0    & 0  & 0 & \\
& -x   & x   & -1    & 1    & 0  & 0 & \\
& -x^2 & x^2 & -2x   & 2x   & 2  & 2 & \\
& -x^3 & x^3 & -3x^2 & 3x^2 & 6x & 6x & \\
& 0   & -1   & 0    & 0    & 0  & 0 & \\
& 0   & -x   & 0    & -1    & 0  & 0 & \\
& 0   & -x^2 & 0    & -2x   & 0  & -2 & \\
& 0   & -x^3 & 0    & -3x^2 & 0  & -6x & \\
\end{bmatrix}$$

So by definition of orthogonal, $u \cdot v = 0$, so if we look at each column of $M^T$ as a column vector, we see that in order for $Mz = 0$, $M^T \cdot z = 0$ as well since the second is equivalent to (for the first column of $M^T$):
$$ M^T \cdot z = \begin{bmatrix} 
1 \\
x  \\
x^2 \\
x^3 \\
-1 \\
-x \\
-x^2 \\
-x^3 \\
0  \\
0   \\
0  \\
0  \\
\end{bmatrix} \cdot
\begin{bmatrix} a_0 \\ b_0 \\ c_0 \\ d_0 \\ a_1 \\ b_1 \\ c_1 \\ d_1 \\ a_2 \\ b_2 \\ c_2 \\ d_2 \end{bmatrix} =
a_0 - a_1 + b_0x - b_1x + c_0x^2 - c_1x^2 + d_0x^3 - d_1x^3 = 0 $$

B. 
```{r}
M.t <- t(M)

A <- cbind(M.t, matrix(rnorm(72), nrow=12, ncol=6))
det(A) #not singular

A.qr <- qr(A)
Q <- qr.Q(A.qr)
Q <- Q[,7:12]


g.x <- function(z = Q[,1], data = male$age) {
  store <- NULL
  for(i in 1:length(data)){
  if(data[i] < 15) {
    s.x1 <- z[1] + z[2]*data[i] + z[3]*data[i]^2 + z[4]*data[i]^3
    store <- c(store, s.x1)
  } else if (data[i] < 20 & data[i] >=15) {
    s.x2 <- z[5] + z[6]*data[i] + z[7]*data[i]^2 + z[8]*data[i]^3
    store <- c(store, s.x2)
  } else if (data[i] >= 20) {
    s.x3 <- z[9] + z[10]*data[i] + z[11]*data[i]^2 + z[12]*data[i]^3
    store <- c(store, s.x3)
  }
  }
  return(store)
}

fit.male <- lapply(1:6, function(x) g.x(z = Q[,x], data = male$age))
```

C. Argue that any cubic spline S(x) with knots ζ1, ζ2 must be of the form S(x) = \Sigma j=1 αjgj (x).
DO THIS LATER

c. Don't know what's wrong here. I think the g basis has problems.
```{r}
h <- list(
h.1 = function(x) return(1),
h.2 = function(x) return(x),
h.3 = function(x) return(x^2),
h.4 = function(x) return(x^3),
h.5 = function(x) return(ifelse((x - 15)^3 > 0, (x - 15)^3, 0)),
h.6 = function(x) return(ifelse((x - 20)^3 > 0, (x - 20)^3, 0))
)

h.i <- lapply(1:6, function(y) h[[y]](x=male$age))

h.plots <- lapply(1:6, function(x) qplot(male$age,unlist(h.i[[x]]), geom="line") + ylab(x))

g.plots <- lapply(1:6, function(x) qplot(male$age,unlist(fit.male[[x]]), geom="line") + ylab(x))

plots <- append(h.plots, g.plots)

multiplot(plotlist = plots, layout=matrix(c(1:12), nrow=6, ncol=2))
```

d. Given $min \Sigma_{i=1}^N (y_i - \Sigma_{j=1}^6 \alpha_j h_j(x_i))^2$; we can express the $h_j(x_i)$ in the form of a matrix W:

$$W = \begin{bmatrix} 
h_1(x_1) & h_2(x_1) & h_3(x_1) & h_4(x_1) & h_5(x_1) & h_6(x_1) \\
h_1(x_2) & h_2(x_2) & h_3(x_2) & h_4(x_2) & h_5(x_2) & h_6(x_2) \\
&&& \vdots &&&\\
h_1(x_n) & h_2(x_n) & h_3(x_n) & h_4(x_n) & h_5(x_n) & h_6(x_n) \\
\end{bmatrix}$$

Thus we can rewrite the minimization function as $min_{\alpha \in \mathbb{R}^6} || y - W\alpha ||^2$.

```{r}
y.male <- male$spnbmd
h.i <- lapply(1:6, function(y) h[[y]](x=male$age))
W.male <- cbind(unlist(h.i[[1]]), unlist(h.i[[2]]), unlist(h.i[[3]]), unlist(h.i[[4]]), unlist(h.i[[5]]), unlist(h.i[[6]])) 

alpha.male <- solve(t(W.male)%*%W.male) %*% t(W.male) %*% y.male 

fit.hm <- W.male %*% alpha.male
qplot(male$age, y.male) + geom_line(aes(y=fit.hm, color="red"))

y.female <- female$spnbmd
h.i <- lapply(1:6, function(y) h[[y]](x=female$age))
W.female <- cbind(unlist(h.i[[1]]), unlist(h.i[[2]]), unlist(h.i[[3]]), unlist(h.i[[4]]), unlist(h.i[[5]]), unlist(h.i[[6]])) 

alpha.female <- solve(t(W.female)%*%W.female) %*% t(W.female) %*% y.female 

fit.hf <- W.female %*% alpha.female
qplot(female$age, y.female) + geom_line(aes(y=fit.hf, color=3))
```

e.
```{r}
y.male <- male$spnbmd
g.i <- lapply(1:6, function(x) g.x(z = Q[,x], data = male$age))
W.gm <- cbind(unlist(g.i[[1]]), unlist(g.i[[2]]), unlist(g.i[[3]]), unlist(g.i[[4]]), unlist(g.i[[5]]), unlist(g.i[[6]])) 

alpha.gm <- solve(t(W.gm)%*%W.gm) %*% t(W.gm) %*% y.male 

fit.gm <- W.gm %*% alpha.gm

qplot(male$age, y.male) + geom_line(aes(y=fit.gm, color="red", size=.3)) + geom_line(aes(y=fit.hm, color="green", size = .1)) + theme(legend.position="none")


y.female <- female$spnbmd
g.i <- lapply(1:6, function(x) g.x(z = Q[,x], data = female$age))
W.gf <- cbind(unlist(g.i[[1]]), unlist(g.i[[2]]), unlist(g.i[[3]]), unlist(g.i[[4]]), unlist(g.i[[5]]), unlist(g.i[[6]])) 

alpha.gf <- solve(t(W.gf) %*% W.gf) %*% t(W.gf) %*% y.female 

fit.gf <- W.gf %*% alpha.gf

qplot(female$age, y.female) + geom_line(aes(y=fit.gf, color="blue", size=.1)) + geom_line(aes(y=fit.hf, size=.01)) + theme(legend.position="none")
```